Style transfer with large language models

### 2025-05-05 paper/main.tex
paper updates 2

% \documentclass[twocolumn]{article}
\documentclass{article}
\usepackage{graphicx}
% \usepackage[T1]{fontenc}
% \usepackage{charter}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{array}
\usepackage{amsmath}
% \usepackage{float}

@@

The rapid advancement of large language models (LLMs) has revolutionized natural language processing (NLP). Although large language models (LLMs) have been used anecdotally for style transfer, they typically do not explicitly disentangle style from content, leading to possible unintended alterations in meaning. Our project aims to bridge this gap by investigating the layers in which style-specific attributes are most effectively represented, allowing independent manipulation of style and content.

\section{Related Work}
Early approaches to text style transfer aimed to disentangle style, drawing inspiration from analogous techniques in image style transfer \citep{gatys2016image, zhu2017unpaired} and subsequently adapted for NLP applications \citep{john2019disentangled, patel2022low}. These methods demonstrated that fine-tuned pre-trained LMs can effectively induce stylistic transformations even with non-parallel data. However, these approaches often treat style and content as inherently intertwined, making it challenging to isolate and control individual attributes.

@@

Building on it, our work adopts a more targeted strategy. By applying linear probes, the present study directly measures how formality and domain-specific features are localized across the layers of an LLM. Such an approach offers the potential to design layer-specific interventions that enhance style transfer performance while preserving content integrity.

@@

These datasets are suitable for our project due to their size and the nature of the style transfer tasks. Additionally, we will also evaluate the ability to maintain content for downstream tasks, such as:

\begin{itemize}
    \item \href{https://ai.google.com/research/NaturalQuestions}{\textit{Natural Questions}}: Question answering (QA) dataset of +300K QA pairs. Each example has query text, and long- and short-forms answers.
    \item \href{https://github.com/abisee/cnn-dailymail}{\textit{CNN/Daily Mail}}: Summarization dataset of $\approx$300K articles paired with human-written summaries.
\end{itemize}

@@

Our methodology involves the following steps:
\begin{enumerate}
    \item \textbf{Dataset Preparation:} Text samples are collected from \href{https://huggingface.co/datasets/osyvokon/pavlick-formality-scores}{Pavlick Formality Scores dataset}. Text samples are preprocessed and split into training and testing sets. In addition, data augmentation strategies are employed to create paired formal-informal examples and avoid content discrepancies.
    \item \textbf{Feature Extraction:} The Phi-4 language model is used to extract representations from different layers by processing text samples. state outputs from each layer are recorded to capture a hierarchy of linguistic features.
    \item \textbf{Linear Probing:} Logistic regression classifiers are trained on the extracted features to predict formality and domain labels.
    \item \textbf{Evaluation:}     Evaluation is conducted on two fronts:\\
    \textbf{Style Classification:}
        The accuracy of the classifiers is measured on both the training and testing sets.

    \textbf{Content Preservation:}
        Two downstream tasks assess how well the underlying semantic content is maintained after style modifications:
        \begin{itemize}
            \item For the question answering task using the Natural Questions dataset, content preservation is evaluated by the \textbf{F1 score}. In this context, Precision is defined as the ratio of correctly predicted answer tokens to the total predicted tokens, while Recall is the ratio of correctly predicted tokens to the total tokens in the reference answer. The F1 score, as the harmonic mean of precision and recall, captures the balance between these measures.
            \item For the summarization task using the CNN/Daily Mail dataset, content preservation is measured using the \textbf{ROUGE-1 metric}. ROUGE-1 computes the unigram overlap between the generated summary and the reference summary, indicating the degree of content similarity.
        \end{itemize}
\end{enumerate}

@@

\begin{itemize}
    \item \textbf{F1 Score (Natural Questions):}
        \begin{equation}
            \text{Precision} = \frac{TP}{TP + FP}
        \end{equation}
        \begin{equation}
            \text{Recall} = \frac{TP}{TP + FN}
        \end{equation}
        \begin{equation}
            \text{$F_1$} = 2 \cdot \frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}
        \end{equation}

    where $TP$, $TN$, $FP$, and $FN$ denote true positives, true negatives, false positives, and false negatives, respectively.

    \item \textbf{ROUGE-1 (CNN/Daily Mail):}

        {\small
        \begin{equation}
            \text{ROUGE-1} =
            \frac{
              \sum_{w \in \text{Ref}}
              \min\!\bigl(\text{Count}_{\text{match}}(w), \text{Count}_{\text{cand}}(w)\bigr)
            }{
              \sum_{w \in \text{Ref}}
              \text{Count}(w)
            }
        \end{equation}
        \normalsize}

\end{itemize}

These metrics ensure that, while stylistic attributes are accurately classified, the semantic content remains preserved following style modifications.

@@

Figure~\ref{fig:formality-accuracy} displays the accuracy plot for formality probing on the Pavlick Formality Dataset. The results indicate that the intermediate layers (layer 13) of the model achieve the highest accuracy, suggesting that these layers are most effective at capturing formality-related features.

\begin{figure*}
    \centering
    \includegraphics[width=1.0\textwidth]{figures/1_formality.png}
    \caption{Formality Probing Accuracy Across Phi-4 Layers}
    \label{fig:formality-accuracy}
\end{figure*}

Figure~\ref{fig:paired-formality-accuracy} presents the accuracy plot for paired formality probing using the Generated Formality Dataset. In this experiment, paired formal-informal examples are used to assess the model’s capability to distinguish between the two styles. Consistent with the single-sample analysis, the intermediate layers (layer 17) again demonstrate superior performance.

\begin{figure*}
    \centering
    \includegraphics[width=1.0\textwidth]{figures/2_pairedFormality.png}
    \caption{Paired Formality Probing Accuracy Across Phi-4 Layers}
    \label{fig:paired-formality-accuracy}
\end{figure*}

\subsection{Domain Probing}
Figure~\ref{fig:domain-accuracy} shows the accuracy plot for domain probing on the Pavlick Formality Dataset. The findings reveal that earlier layers (layer 8) exhibit higher predictive accuracy for domain-specific attributes, further confirming that stylistic and domain information are localized in particular regions of the model.

\begin{figure*}
    \centering
    \includegraphics[width=1.0\textwidth]{figures/3_domain.png}
    \caption{Domain Probing Accuracy Across Phi-4 Layers}
    \label{fig:domain-accuracy}
\end{figure*}

@@

\subsubsection{Training Process}
The model was trained on the Pavlick Formality Dataset for 20 epochs (Fig.~\ref{fig:training-loss}) using the Adam optimizer with a learning rate of 1e-4.
We employed a weighted cross-entropy loss to account for class imbalance:

@@

\subsubsection{Domain Classification Performance}
Figure~\ref{fig:domain-confusion} presents the confusion matrix for domain classification using our style latent approach. The model achieves high accuracy in distinguishing between different domains, with particularly strong performance in identifying content from the EMAILS and NEWS categories.

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/4_confusion.png}
    \caption{Confusion Matrix for Domain Classification using the Style Latent Model}
    \label{fig:domain-confusion}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/training-loss2.png}
    \caption{Training Loss Curve for Style Latent Model.}
    \label{fig:training-loss}
\end{figure}

@@

\subsubsection{Content Preservation}
Content preservation metrics were integrated by evaluating the Natural Questions dataset using the F1 score for question answering, yielding an F1 score of 0.72. Additionally, we evaluated both the baseline and style-modified models on the CNN/Daily Mail dataset using the ROUGE metric for summarization, as shown in Table~\ref{tab:rouge-scores}.

\begin{table*}[ht]
\centering
\caption{ROUGE Score Comparison Between Baseline and Style-Modified Phi-4 Models}
\label{tab:rouge-scores}
\begin{tabular}{lcccc}
\hline
\textbf{Model} & \textbf{ROUGE-1} & \textbf{ROUGE-2} & \textbf{ROUGE-L} & \textbf{ROUGE-Lsum} \\
\hline
Baseline Phi-4 & 0.1518 & 0.0776 & 0.1071 & 0.1253 \\
Modified Phi-4 & 0.1470 & 0.0753 & 0.1049 & 0.1224 \\
Difference  & -0.031 & -0.029 & -0.020 & -0.023 \\
\hline
\end{tabular}
\end{table*}

\begin{table*}
\centering
\caption{Examples of Style Transfer with Different Domain Latents}
\label{tab:style-transfer-examples}
\begin{tabular}{>{\raggedright\arraybackslash}p{1.5cm} >{\raggedright\arraybackslash}p{14cm}}
\hline
\textbf{Style} & \textbf{Generated Text} \\
\hline
Original & The MDC has a long list of complaints it alleges are non-fulfillment of the Global Political Agreement and wants resolved before it can participate fully in government. \\
\hline
ANSWERS & MDC also accused the ZANU PF government of being a one party state and said the MDC is not treated as an equal partner... \\
\hline
BLOG & So, MDC accused the government of failing to address key issues. \\
\hline
EMAIL & the MDC-T says it wants the reserve bank gov to be replaced and that it wants to know what is in the arms deal signed by the government? \\
\hline
\hline
Original & How does it affect dogs? \\
\hline
ANSWERS & What are the symptoms? What should I do if my dog has it? And, can my dog give it to me? \\
\hline
BLOG & Dogs can suffer from various neurological disorders, including epilepsy, which can cause seizures and other symptoms affecting their nervous system. \\
\hline
NEWS & The disease can cause severe symptoms in dogs, including diarrhea, vomiting, and lethargy. It can be particularly dangerous for puppies, elderly dogs, or those with compromised immune systems. \\
\hline
EMAIL & **What are common signs of poisoning in dogs?** - **Answer:** Common signs include vomiting, diarrhea, excessive drooling, lethargy, difficulty breathing, seizures, and loss of coordination. \\
\hline
\end{tabular}
\end{table*}

These results confirm minor compromise in semantic integrity despite style modifications, with only a small reduction (approximately 3\%) in ROUGE scores compared to the baseline model. This minimal degradation in content preservation metrics demonstrates that our style embedding approach effectively separates style representation from content, allowing style manipulation without substantial impact on the underlying meaning.

@@

Table~\ref{tab:style-transfer-examples} illustrates examples of style transfer using our approach, where we applied different domain-specific latent vectors to the same content.
The table showcases how the original text is rendered differently when encoded with ANSWERS, BLOG, and EMAIL style latents.
Note how each variant maintains similar (but clearly imperfect) core information while adopting characteristic stylistic elements of its domain: the ANSWERS style uses more accusatory language, the BLOG style incorporates temporal markers and detailed reporting, and the EMAIL style employs a more direct, abbreviated approach with question formatting.
These examples demonstrate the model's ability to transform text style while preserving the underlying semantic content.

In the second example, notice the confusion where the EMAIL generation (which begins with \textit{What are common signs of poisoning in dogs?}) begins with a question and forms what appears to be an ANSWERS style response, reflecting the findings from the domain probing experiments that EMAIL and ANSWERS are stylistically similar and often confused (Fig.~\ref{fig:domain-confusion}).

>>>>

\documentclass[twocolumn]{article}
% \documentclass{article}
\usepackage{graphicx}
% \usepackage[T1]{fontenc}
% \usepackage{charter}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{array}
\usepackage{amsmath}
\usepackage{float}
\usepackage{tabularx}

@@

The rapid advancement of large language models (LLMs) has revolutionized natural language processing (NLP). Although large language models (LLMs) have been used anecdotally for style transfer, they typically do not explicitly disentangle style from content, leading to possible unintended alterations in meaning. Our project aims to bridge this gap by investigating the layers in which style-specific attributes are most effectively represented, allowing independent manipulation of style and content. Furthermore, we aim to modify these layers to determine if tweaking them can produce results that align more closely with the desired style.

\section{Related Work}
Early approaches to text style transfer aimed to disentangle style, drawing inspiration from analogous techniques in image style transfer \citep{gatys2016image, zhu2017unpaired} and subsequently adapted for NLP applications \citep{john2019disentangled, patel2022low}. These methods demonstrated that fine-tuned pre-trained LLMs can effectively induce stylistic transformations even with non-parallel data. However, these approaches often treat style and content as inherently intertwined, making it challenging to isolate and control individual attributes.

@@

Building on this, our work adopts a more targeted strategy. By applying linear probes, we directly measure how formality and domain-specific features are localized across the layers of an LLM. Based on these findings, we design layer-specific interventions targeting the most style-aware sections of the model to enhance style transfer performance while preserving content integrity.

@@

These datasets are suitable for our project due to their size and the nature of the style transfer tasks. Additionally, we will also evaluate the ability to maintain content for downstream tasks with:

\begin{itemize}
    \item \href{https://github.com/abisee/cnn-dailymail}{\textit{CNN/Daily Mail}}: Summarization dataset of $\approx$300K articles paired with human-written summaries.
\end{itemize}

@@

Our methodology involves the following steps:
\begin{enumerate}
    \item \textbf{Dataset Preparation:} Text samples are collected from Pavlick Formality Scores dataset. Text samples are preprocessed and split into training and testing sets. In addition, data augmentation strategies are employed to create paired formal-informal examples to avoid content discrepancies.
    \item \textbf{Feature Extraction:} The Phi-4 language model is used to extract representations from different layers by processing text samples. State outputs from each layer are recorded to capture a hierarchy of linguistic features.
    \item \textbf{Linear Probing:} Logistic regression classifiers are trained on the extracted features to predict formality and domain labels.
    \item \textbf{Evaluation:}     Evaluation is conducted on two fronts:\\
    \textbf{Style Classification:}
        The accuracy of the classifiers is measured on both the training and testing sets.

    \textbf{Content Preservation:}
        To assess how well the underlying semantic content is maintained after style modifications, content preservation is measured using:
        \begin{itemize}
            \item[--] \textbf{BLEU}: BLEU-1, BLEU-2, BLEU-3, and BLEU-4 are used to evaluate the n-gram overlap between the generated summary and the reference summary. These metrics provide a quantitative measure of how well the generated text aligns with the original content.
            \item[--] \textbf{ROUGE}: ROUGE-N computes the overlap between the generated summary and the reference summary, indicating the degree of content similarity. ROUGE-L evaluates the length of the longest common subsequence (LCS) between the candidate and reference, emphasizing sentence-level structure and fluency. ROUGE-Lsum extends ROUGE-L by computing LCS-based similarity across sentences and aggregating the result, offering a more robust evaluation for multi-sentence or document-level summaries.
        \end{itemize}
\end{enumerate}

@@

\begin{itemize}
    \item \textbf{BLEU-N:}
        \begin{equation}
            \text{BLEU-N} = BP \cdot \exp\left(\sum_{n=1}^{N} w_n \cdot \log p_n\right)
        \end{equation}
        Where $BP$ is the brevity penalty, $p_n$ is the precision of n-grams, and $w_n$ is the weight for each n-gram.
        \begin{equation}
            \text{BP} =
            \begin{cases}
                1 & \text{if } c > r \\
                e^{(1 - \frac{r}{c})} & \text{if } c \leq r
            \end{cases}
        \end{equation}
        Here, $c$ is the length of the candidate summary, and $r$ is the reference length.
    \item \textbf{ROUGE-N:}
        \small
            \begin{equation}
                \frac{
                \sum_{g \in \text{Ref}_N} \min\left(\text{Count}_{\text{match}}(g), \text{Count}_{\text{cand}}(g)\right)
                }{
                \sum_{g \in \text{Ref}_N} \text{Count}(g)
                }
            \end{equation}
        \normalsize
        Where $g$ is an n-gram and $\text{Ref}_N$ denotes all reference n-grams of order $N$.            
    \item \textbf{ROUGE-L:}
        \begin{equation}
        R_{LCS} = \frac{\text{LCS}(X, Y)}{\text{length}(Y)}, \quad
        P_{LCS} = \frac{\text{LCS}(X, Y)}{\text{length}(X)}
        \end{equation}
        \begin{equation}
        \text{ROUGE-L} = \frac{(1 + \beta^2) \cdot R_{LCS} \cdot P_{LCS}}{R_{LCS} + \beta^2 \cdot P_{LCS}}
        \end{equation}
        Where $\text{LCS}(X, Y)$ is the length of the longest common subsequence between candidate $X$ and reference $Y$. $\beta = 1$ to equally weight precision and recall.
\end{itemize}

@@

Figure~\ref{fig:formality-accuracy} displays the accuracy plot for formality probing on the Pavlick Formality Dataset. The results indicate that the intermediate layers (layer 13) of the model achieve the highest accuracy, suggesting that these layers are most effective at capturing formality-related features.
Figure~\ref{fig:paired-formality-accuracy} presents the accuracy plot for paired formality probing using the Generated Formality Dataset. In this experiment, paired formal-informal examples are used to assess the model’s capability to distinguish between the two styles. Consistent with the single-sample analysis, the intermediate layers (layer 17) again demonstrate superior performance.
\subsection{Domain Probing}
Figure~\ref{fig:domain-accuracy} shows the accuracy plot for domain probing on the Pavlick Formality Dataset. The findings reveal that domain-specific attributes are represented across all layers of the model, with no single layer being exclusively responsible for selectively encoding domain information. It further confirms that stylistic and domain information are not localized in particular layers but are distributed across the model.

\onecolumn
\begin{figure*}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/1_formality.png}
    \caption{Formality Probing Accuracy Across Phi-4 Layers}
    \label{fig:formality-accuracy}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/2_pairedFormality.png}
    \caption{Paired Formality Probing Accuracy Across Phi-4 Layers}
    \label{fig:paired-formality-accuracy}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/3_domain.png}
    \caption{Domain Probing Accuracy Across Phi-4 Layers}
    \label{fig:domain-accuracy}
\end{figure*}
\twocolumn

\onecolumn

\begin{table*}[ht]
\centering
\caption{Content Preservation Metrics of Baseline and Style-Modified Model}
\label{tab:content-preservation}
\begin{tabular}{lcccc}
\hline
\textbf{Metric} & \textbf{Baseline} & \textbf{Modified} & \textbf{Diff.} \\
\hline
\textbf{BLEU-1} & 0.0767 & 0.0761 & -0.0006 \\
\textbf{BLEU-2} & 0.0503 & 0.0501 & -0.0002 \\
\textbf{BLEU-3} & 0.0353 & 0.0352 & -0.0001 \\
\textbf{BLEU-4} & 0.0249 & 0.0248 & -0.0001 \\
\textbf{ROUGE-1 F1} & 0.1584 & 0.1582 & -0.0002 \\
\textbf{ROUGE-2 F1} & 0.0796 & 0.0795 & -0.0001 \\
\textbf{ROUGE-L F1} & 0.1113 & 0.1111 & -0.0001 \\
\textbf{ROUGE-1 Precision} & 0.0888 & 0.0886 & -0.0002 \\
\textbf{ROUGE-1 Recall} & 0.8897 & 0.8897 & 0.0000 \\
\hline
\end{tabular}
\end{table*}

\begin{table}[H]
    \centering
    \caption{Examples of Style Transfer with Different Domain Latents}
    \label{tab:style-transfer-examples}
    \begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{2.5cm} X}
        \hline
        \textbf{Style} & \textbf{Generated Text} \\
        \hline
        Original & The MDC has a long list of complaints it alleges are non-fulfillment of the Global Political Agreement and wants resolved before it can participate fully in government. \\
        \hline
        ANSWERS & MDC also accused the ZANU PF government of being a one party state and said the MDC is not treated as an equal partner... \\
        \hline
        BLOG & So, MDC accused the government of failing to address key issues. \\
        \hline
        EMAIL & the MDC-T says it wants the reserve bank gov to be replaced and that it wants to know what is in the arms deal signed by the government? \\
        \hline
        \hline
        Original & How does it affect dogs? \\
        \hline
        ANSWERS & What are the symptoms? What should I do if my dog has it? And, can my dog give it to me? \\
        \hline
        BLOG & Dogs can suffer from various neurological disorders, including epilepsy, which can cause seizures and other symptoms affecting their nervous system. \\
        \hline
        NEWS & The disease can cause severe symptoms in dogs, including diarrhea, vomiting, and lethargy. It can be particularly dangerous for puppies, elderly dogs, or those with compromised immune systems. \\
        \hline
        EMAIL & **What are common signs of poisoning in dogs?** - **Answer:** Common signs include vomiting, diarrhea, excessive drooling, lethargy, difficulty breathing, seizures, and loss of coordination. \\
        \hline
    \end{tabularx}
\end{table}
\twocolumn

@@

\subsubsection{Training Process}
The model was trained on the Pavlick Formality Dataset for 20 epochs (Fig.~\ref{fig:training-loss}) using the Adam optimizer with a learning rate of 1e-4.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.5\linewidth]{figures/training-loss2.png}
    \caption{Training Loss Curve for Style Latent Model.}
    \label{fig:training-loss}
\end{figure}

We employed a weighted cross-entropy loss to account for class imbalance:

@@

\subsubsection{Domain Classification Performance}
Figure~\ref{fig:domain-confusion} presents the confusion matrix for domain classification using our style latent approach. The model achieves $accuracy=0.54$, being able to distinguish between different domains, with particularly strong performance in identifying content from the EMAILS and NEWS categories.

\begin{figure} [H]
    \centering
    \includegraphics[width=0.8\linewidth]{figures/4_confusion.png}
    \caption{Confusion Matrix for Domain Classification using the Style Latent Model}
    \label{fig:domain-confusion}
\end{figure}

@@

\subsubsection{Content Preservation}
Content preservation metrics were integrated by evaluating both the baseline and style-modified models on the CNN/Daily Mail dataset, as shown in Table~\ref{tab:content-preservation}.

A negligible reduction (\(<0.01\%\)) in BLUE and ROUGE scores was observed for the style-modified model compared to the baseline, confirming no compromise in semantic integrity despite style modifications. This minimal degradation in content preservation metrics demonstrates that our style embedding approach effectively separates style representation from content, allowing style manipulation without substantial impact on the underlying meaning.

@@

Table~\ref{tab:style-transfer-examples} illustrates examples of style transfer using our approach, where we applied different domain-specific latent vectors to the same content.
The table showcases how the original text is rendered differently when encoded with ANSWERS, BLOG, and EMAIL style latents.
Note how each variant maintains similar (but clearly imperfect) core information while adopting characteristic stylistic elements of its domain: the ANSWERS style uses more accusatory language, the BLOG style incorporates temporal markers and detailed reporting, and the EMAIL style employs a more direct, abbreviated approach with question formatting.
These examples demonstrate the model's ability to transform text style while preserving the underlying semantic content.
In the second example, notice the confusion where the EMAIL generation (which begins with \textit{What are common signs of poisoning in dogs?}) begins with a question and forms what appears to be an ANSWERS style response, reflecting the findings from the domain probing experiments that EMAIL and ANSWERS are stylistically similar and often confused (Fig.~\ref{fig:domain-confusion}).
